---
title: "ST-558, Project 2"
author: "Nataliya Peshekhodko"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Vignette for reading and summarizing data from an IQAir APIs

## Overall Goal

Vignettes are explanations of some concept, package, etc. with text, code, and output interweaved. Our goal with this project is to create a vignette about contacting an [IQAir APIs](https://www.iqair.com/dashboard/api) using custom functions to query, parse, and return well-structured data. Once functions are available, basic exploratory data analysis is performed on data returned by API endpoints. IQAir APIs provide information about **real time air quality data and weather metrics**.

## Packages

Packages required to run code included in that analysis:

```{r, message=FALSE, warning=FALSE}
library(httr)
library(dplyr)
library(tibble)
library(jsonlite)
library(ggplot2)
library(RColorBrewer)
```


  - `httr` - required for making HTTP requests
  - `dplyr` - required for data manipulation
  - `tibble` - required for formatting data frame
  - `jsonlite` - required for converting JSON data from/to R objects
  - `ggplot2` - required for creating graphs
  - `RColorBrewer` - provides color schemes for maps
  

Define variable for API key required for the authentication during HTTP request:

```{r}
key = '97d5c2b8-7656-4974-827e-9d59486f7777'
```


## Functions for reading data from IQAir APIs

This section is dedicated for functions developed for `IQAir APIs` interaction. 


### Function to get states

Function to retrieve `states` based on the `country` name. States returned as a list.


```{r}
get_states <- function(country, api_key = key) {
  url = 'http://api.airvisual.com/v2/states'
  resp = GET(url, query = list(key = api_key, 
                               country = country))
  parsed = fromJSON(rawToChar(resp$content))
  
  return (as.list(parsed$data)$state)
}
```


### Function to get cities

Function to get `cities` based on the provided `country` and `state`. Cities returned as a list.


```{r}
get_cities <- function(country, state, api_key = key) {
  url = 'http://api.airvisual.com/v2/cities'
  
  resp = GET(url, query = list(key = api_key, 
                               country = country,
                               state = state))
  parsed = fromJSON(rawToChar(resp$content))
  return (as.list(parsed$data)$city)
}
```


### Function Get air quality and weather metrics based on country, state and city

Function returns air quality based on the `country`, `state` and `city`. 

If `city` is not provided, random sample of the size `sample_size` will be chosen from the cities which belong to the `state`. 


If `weather_metrics = TRUE` in addition to air quality metrics, weather metrics like, temperature, humidity, wind speed and atmospheric pressure will be returned. 

Function returns data frame in `tibble` format with following columns:

  - `country`
  - `state`
  - `city`
  - `aqius` - AQI value based on US EPA standard
  - `aqicn` - AQI value based on China MEP standard

If  `weather_metrics = TRUE` in addition to the columns above, function also returns:


  - `temp_cels` - temperature in Celsius
  - `humidity` - humidity %
  - `wind_speed` - wind speed (m/s)
  - `atm_pressure` - atmospheric pressure in hPa


```{r}
get_air_quality_per_city <- function(country, 
                                     state, 
                                     city = NULL,
                                     weather_metrics = TRUE,
                                     sample_size = 5,
                                     api_key = key) {
  
  results = tibble()
  url = 'http://api.airvisual.com/v2/city'
 
  if (is.null(city)) {
    cities = get_cities(country, state)
  
    if (length(cities) > sample_size) {
      set.seed(123)
      cities = sample(cities, size = sample_size)
    }
    
    for (city in cities) {
      Sys.sleep(15) # required to avoid error - too many requests
      resp = GET(url, query = list(key = key, 
                                   country = country,
                                   state = state,
                                   city = city))
      parsed = fromJSON(rawToChar(resp$content))
      
      if (weather_metrics == TRUE) {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country,
                           temp_cels = parsed$data$current$weather$tp,
                           humidity = parsed$data$current$weather$hu, 
                           wind_speed = parsed$data$current$weather$ws,
                           atm_pressure = parsed$data$current$weather$pr)
      } else {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country)
      }
      results <- bind_rows(results, subset_df)
    
  }
    } else {
      resp = GET(url, query = list(key = key, 
                                   country = country,
                                   state = state,
                                   city = city))
      parsed = fromJSON(rawToChar(resp$content))
      if (weather_metrics == TRUE) {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country,
                           temp_cels = parsed$data$current$weather$tp,
                           humidity = parsed$data$current$weather$hu, 
                           wind_speed = parsed$data$current$weather$ws,
                           atm_pressure = parsed$data$current$weather$pr)
      } else {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country)
      }
      results <- bind_rows(results, subset_df)
  }
  return (results)
}
```


### Function to get air quality and weather metrics for country

Function returns air quality and weather metrics based on the country name. Random sample of the size `sample_size` is thrown from from the states which belongs to the country. For each state, random city is selected and metrics returned.


Functions returns data frame in `tibble` format with following columns:

  - `country`
  - `state`
  - `city`
  - `aqius` - AQI value based on US EPA standard
  - `aqicn` - AQI value based on China MEP standard
  - `temp_cels` - temperature in Celsius
  - `humidity` - humidity %
  - `wind_speed` - wind speed (m/s)
  - `atm_pressure` - atmospheric pressure in hPa


```{r}
get_air_quality_per_country <- function(country, 
                                        weather_metrics = TRUE,
                                        sample_size = 5,
                                        api_key = key) {
  
  results = tibble()
  url = 'http://api.airvisual.com/v2/city'
  
  
  states = get_states(country, api_key = key)
  
  set.seed(2)
  states = sample(states, size = sample_size)
  
  
  for (state in states) {
    Sys.sleep(15) # required to avoid error - too many requests
    cities = get_cities(country, state, api_key = key)
    set.seed(4)
    city = sample(cities, size = 1)
    
    Sys.sleep(15) # required to avoid error - too many requests
    resp = GET(url, query = list(key = api_key, 
                                 country = country,
                                 state = state,
                                 city = city))
    parsed = fromJSON(rawToChar(resp$content))
    subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                       aqicn = parsed$data$current$pollution$aqicn, 
                       city = parsed$data$city,
                       state = parsed$data$state,
                       country = parsed$data$country,
                       temp_cels = parsed$data$current$weather$tp,
                       humidity = parsed$data$current$weather$hu, 
                       wind_speed = parsed$data$current$weather$ws,
                       atm_pressure = parsed$data$current$weather$pr)
    results <- bind_rows(results, subset_df)
  }
  return (results)
}
```


## Exploratory Data Analysis

EDA analysis will be done on countries level and, for the United States, on states level.

### EDA for countries


Let's look at air quality and weather metrics for different countries across the globe. Chosen countries: **China**, **USA**, **Poland**, **Australia** and **Norway**.

*Note: Sample size and set of countries chosen to be small due to API limitations and time required to render the whole document* 

```{r, cache=TRUE}
sample = 5
air_usa = get_air_quality_per_country(country = 'USA', sample_size = sample)
air_china = get_air_quality_per_country(country = 'China', sample_size = sample)
air_poland = get_air_quality_per_country(country = 'Poland', sample_size = sample)
air_australia = get_air_quality_per_country(country = 'Australia', sample_size = sample)
air_norway = get_air_quality_per_country(country = 'Norway', sample_size = sample)
```

Combine all obtained data frames together in one data frame.

```{r}
all = bind_rows(air_australia, air_poland, air_china, air_usa, air_norway)
```

Let's look at data frame with combined records:
```{r}
all
```

Now, we need to create categorical variables for `aqius`, `aqicn` and `humidity`. 

If `aqius` value is less than **50**, the air quality is considered to be `good`. If `aqius` value is greater or equal **50** and less or equal **100**,  the air quality is considered to be `moderate`. If `aquis` value is greater than **100**, the air quality is considered `unhealthy`.

If `aqicn` value is less than **50**, the air quality is considered to be `excellent`, if `aqicn` value is greater or equal **50** and lees or equal **100**, the air quality is considered to be `good`.  If `aqicn` value is greater than **100**, the air quality is considered `polluted`.

If `humidity` level is less or equal **40%**, it is considered `low`. If `humidity` level is greater than **40%** and less than or equal to **60%**, it is considered `normal`. If `humidity` is greater than **60%**, it is considered `high`.

Creating new categorical variables based on the rules outlined above and displaying updated data frame.

```{r}
all <- all %>%
  mutate(aqius_category = case_when(
    aqius < 50 ~ "good",
    aqius >= 50 & aqius <= 100 ~ "moderate",
    aqius > 100 ~ "unhealthy"
  )) %>%
  mutate(aqicn_category = case_when(
    aqicn < 50 ~ "excellent",
    aqicn >= 50 & aqicn <= 100 ~ "good",
    aqicn > 100 ~ "polluted"
  )) %>%
  mutate(humidity_category = case_when(
    humidity <= 40 ~ 'low',
    humidity > 40 & humidity <= 60 ~ 'normal',
    humidity > 60 ~ 'high'
  ))
 all 
```

*Note: in the following sections I will be doing EDA and results interpretation. Given the data, I am retrieving every time I render the document, is real-time data, it could be some discrepancies between observed values and I my interpretation, but I will try my best to be consistent with values and interpretation.*

Create 2-way contingency table for `aqius_category` vs `country`.

```{r}
table(all$aqius_category, all$country)
```
As it could be seen from the table above, `China` has at least one city with `unhealthy` air based on `aqius` values. All sampled cities for `Australia` and `Norway` have `good` air quality. 

Create 2-way contingency table for `aqicn_category` vs `country`.

```{r}
table(all$aqicn_category, all$country)
```

Based on the `aqicn` values, all selected cities for all countries, except `China` have `excellent` air quality.

Create 2-way contingency table for `aqius_category` vs `humidity_category`.

```{r}
table(all$aqius_category, all$humidity_category)
```


Create 2-way contingency table for `aqicn_category` vs `humidity_category`.

```{r}
table(all$aqicn_category, all$humidity_category)
```

Based on the two tables above, the most observations for the `excellent` and `good` air quality are for the `high` humidity level.


Now, let's group data by `country` and calculate `aqius` mean, `aqicn` mean and standard deviations for each group.

```{r}
result <- all %>%
  group_by(country) %>%
  summarize(mean_aqius = mean(aqius), 
            mean_aqicn = mean(aqicn), 
            sd_aquis = sd(aqius),
            sd_aqicn = sd(aqicn)) 
result
```
Based on the values above, `China` has the highest values for `aqius` and `aqicn`, while `Australia` has the lowest values for `aqius` and `aqicn`. 


Visualizing obtained results for `aqius`.

```{r}
ggplot(result, aes(x = country, y = mean_aqius)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.5) +
  labs(title = "Means of AQIUS per country", x = "Country", y = "Mean of AQIUS")
```

Visualizing obtained results for `aqicn`.

```{r}
ggplot(result, aes(x = country, y = mean_aqicn)) +
  geom_bar(stat = "identity", fill = "red", alpha = 0.5) +
  labs(title = "Means of AQICN per country", x = "Country", y = "Mean of AQUCN")
```

The graphs above shows means for the `aqius` and `aqicn` for the selected coutries. `China` has the highest values for `aqius` and `aqicn`, while `Australia` has the lowest values for `aqius` and `aqicn`.

Let's look at `aqius` distribution for each country. Creating box plot for `aqius` per `country`.

```{r}
ggplot(all, aes(x = country, y = aqius, fill = country)) +
  geom_boxplot() +
  labs(title = "AQIUS values distribution per country", x = "Country", y = "AQIUS") +
  theme_minimal()
```
Creating heatmap for numerical variables

```{r}
data <- as.matrix(all %>% select(aqius, aqicn, temp_cels, humidity, wind_speed, atm_pressure))
heatmap(data,
        Rowv=NA,
        Colv=NA, 
        labCol=colnames(data), 
        col= colorRampPalette(brewer.pal(8, "Oranges"))(25), 
        scale="column",
        main="Heatmap for numerical variables")
legend(x="bottomright", legend=c("min", "ave", "max"), 
     fill=colorRampPalette(brewer.pal(8, "Oranges"))(3))
```

Let's looks at correlation matrix for all numerical variables.

```{r}
cor(data)
```

From the matrix above, it could be seen that there are some positive linear correlation between `aqius` and `atm_pressure`, `aqius` and `temp_cels`, `aqicn` and `temp_cels`, `aqicn` and `atm_pressure`.  There are some negative linear correlation between `aqius` and `wind_speed`, `aqicn` and `humidity`, `aqicn` and `wind_speed`.


Let's check if there are any linear dependency between current temperature and observed `aquis` values.

```{r}
ggplot(all, aes(x = temp_cels, y = aqius)) +
  geom_point() +          
  geom_smooth(method = "lm",
              formula = y ~ x,
              se = TRUE) +
  labs(title = "Temperature vs AQIUS",
       x = "Temperature, Celsius",
       y = "AQIUS")  
```

As, it could be seen from the graph above, there are might be some relationship between temperature and observed `aquis` values. The higher the temperature, the higher the `aquis` values.



### EDA for the US states

Now, let's look at several states across `USA` and compare air quality in the context . Chosen states: *California*, *Colorado*, *Minnesota*, *Florida* and *Michigan*. Sample size for each state is equal to `5`.


Reading data for each state using function `get_air_quality_per_city`:

```{r, cache=TRUE}
sample = 5
air_california = get_air_quality_per_city(country = 'USA', state = 'California', sample_size = sample)
air_colorado = get_air_quality_per_city(country = 'USA', state = 'Colorado', sample_size = sample)
air_minnesota = get_air_quality_per_city(country = 'USA', state = 'Minnesota', sample_size = sample)
air_florida = get_air_quality_per_city(country = 'USA', state = 'Florida', sample_size = sample)
air_michigan = get_air_quality_per_city(country = 'USA', state = 'Michigan', sample_size = sample)
```

Combine results in one data frame:

```{r}
all_per_state = bind_rows(air_michigan, air_florida, air_minnesota, air_colorado, air_california)
all_per_state
```


Now, let's use the same rules as outlined above, to create categorical variables `aqius_category`, `aqicn_category` and `humidity_category` and display updated data frame.

```{r}
all_per_state <- all_per_state %>%
  mutate(aqius_category = case_when(
    aqius < 50 ~ "good",
    aqius >= 50 & aqius <= 100 ~ "moderate",
    aqius > 100 ~ "unhealthy"
  )) %>%
  mutate(aqicn_category = case_when(
    aqius < 50 ~ "excellent",
    aqius >= 50 & aqius <= 100 ~ "good",
    aqius > 100 ~ "polluted"
  )) %>%
  mutate(humidity_category = case_when(
    humidity <= 40 ~ 'low',
    humidity > 40 & humidity <= 60 ~ 'normal',
    humidity > 60 ~ 'high'
  ))
all_per_state 
```


Create 2-way contingency table for `aqius_category` vs `state`.


```{r}
table(all_per_state$aqius_category, all_per_state$state)
```

Based on the table above, Michigan has 4 states out of 5 with `moderate` air quality. California, Colorado and Florida have all 5 cities with `good` air quality. 

Create 2-way contingency table for `aqicn_category` vs `state`.

```{r}
table(all_per_state$aqicn_category, all_per_state$state)
```
Based on the table above, Michigan has 4 states out of 5 with `good` air quality.


Create 2-way contingency table for `aqius_category` vs `humidity_category`.


```{r}
table(all_per_state$aqius_category, all_per_state$humidity_category)
```
Most of the observations with `good` air quality are from observations with `high` humidity. 

Create 2-way contingency table for `aqicn_category` vs `humidity_category`.


```{r}
table(all_per_state$aqicn_category, all_per_state$humidity_category)
```
Most of the observations with `exellent` air quality are from observations with `high` humidity. 

Let's looks at correlation matrix for all numerical variables.

```{r}
data <- as.matrix(all_per_state %>% select(aqius, aqicn, temp_cels, humidity, wind_speed, atm_pressure))
cor(data)
```

From the matrix above, it could be seen that there are some positive linear correlation between `aqius` and `atm_pressure`, `aqius` and `temp_cels`, `aqius` and `humidity`, `aqicn` and `temp_cels`, `aqicn` and `atm_pressure`, `aqicn` and `humidity`.  There are some negative linear correlation between `aqius` and `wind_speed`, `aqicn` and `wind_speed`.


Now, let's group by `state` and calculate `aqius` mean, `aqicn` mean and standard deviations for each group.

```{r}
result <- all_per_state %>%
  group_by(state) %>%
  summarize(mean_aqius = mean(aqius), 
            mean_aqicn = mean(aqicn), 
            sd_aquis = sd(aqius),
            sd_aqicn = sd(aqicn)) 
result
```



Let's look at linear relationship between wind speed and observed `aquis` values.


```{r}
ggplot(all_per_state, aes(x = wind_speed, y = aqius)) +
  geom_point() +                 
  geom_smooth(method = "lm", 
              formula = y ~ x,
              se = TRUE) + 
  labs(title = "Temperature vs AQIUS", 
       x = "Wind speed",
       y = "AQIUS")  
```

  As it could be seen from the graph, the stronger the wind, the lower the `aqius` values. 


Let's check if there are any linear dependency between temperature and observed `aquis` values.  

```{r}
ggplot(all_per_state, aes(x = temp_cels, y = aqius)) +
  geom_point() + 
  geom_smooth(method = "lm", 
              formula = y ~ x,
              se = TRUE) + 
  labs(title = "Temperature vs AQIUS", 
       x = "Temperature", 
       y = "AQIUS")  
```

As it could be seen from the graph above, the higher the temperature, the higher `aqius` values.


Let's check if there are any linear dependency between humidity level and observed `aquis` values. 

```{r}
ggplot(all_per_state, aes(x = humidity, y = aqius)) +
  geom_point() + 
  geom_smooth(method = "lm",
              formula = y ~ x, 
              se = TRUE) + 
  labs(title = "Humidity vs AQIUS", 
       x = "Humidity",
       y = "AQIUS") 
```

As it could be seen from the graph above, there is may be some linear relationship between humidity levels and `aqius` values. The higher the humidity, the higher the `aquis` values.


Let's check if there are any linear dependency between atmospheric pressure and observed `aquis` values. 

```{r}
ggplot(all_per_state, aes(x = atm_pressure, y = aqius)) +
  geom_point() + 
  geom_smooth(method = "lm",
              formula = y ~ x, 
              se = TRUE) + 
  labs(title = "Atmospheric pressure vs AQIUS", 
       x = "Atmospheric pressure in hPa",
       y = "AQIUS") 
```

As it could be seen from the graph above, the higher the atmospheric pressure, the higher the `aqius`.


Let's look at `aqius` distribution for each `state`. Creating box plot for `aqius` per `state`.

```{r}
ggplot(all_per_state, aes(x = state, y = aqius, fill = state)) +
  geom_boxplot() +
  labs(title = "AQIUS per state", x = "State", y = "AQIUS") +
  theme_minimal()
```

## Summary

At this project, functions to interact with **IQAir APIs** were created. Functions allow to retrieve data quality and weather metrics per country, state and city. Due to free account limitations, small samples were derived for several countries across the globe and several states across the United States.  For both data sets EDA was conducted. The EDA showed that:

  - there are negative linear correlation between wind speed and air quality
  - air quality is better when humidity level is high
  - air quality is decreasing when atmospheric pressure is increasing.
  
**IQAir APIs** are more suitable to retrieve granular information per city. In order to retrieve large sample of the data, many API calls should be done. 
  


