---
title: "ST-558  Project 2 (Reading and summarizing data from an IQAir APIs)"
author: "Nataliya Peshekhodko"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overall Goal

Vignettes are explanations of concepts, packages, or other topics that interweave text, code, and output. Our project's goal is to create a vignette focused on interacting with the [IQAir APIs](https://www.iqair.com/dashboard/api) using custom functions to query, parse, and retrieve well-structured data. After the functions are available, we conduct basic exploratory data analysis on the data obtained from API endpoints. 


The **IQAir APIs** provide information about **real-time air quality** and **weather metrics**.

# Packages

Packages required to run code included in that analysis:

```{r, message=FALSE, warning=FALSE}
library(httr)
library(dplyr)
library(tibble)
library(jsonlite)
library(ggplot2)
library(RColorBrewer)
library(corrplot)
```


  - `httr` - required for making HTTP requests
  - `dplyr` - required for data manipulation
  - `tibble` - required for formatting data frame
  - `jsonlite` - required for converting JSON data from/to R objects
  - `ggplot2` - required for creating graphs
  - `RColorBrewer` - provides color schemes for maps
  - `corrplot` - required for correlation matrix visualizing
  


# Functions for reading data from IQAir APIs

This section is dedicated for functions developed for `IQAir APIs` interaction. 

Create variable with API key required for the authentication during HTTP request:

```{r}
key = '97d5c2b8-7656-4974-827e-9d59486f7777'
```


## Function to get states

Function to retrieve `states` based on the `country` name. States returned as a list.


```{r}
get_states <- function(country, api_key = key) {
  url = 'http://api.airvisual.com/v2/states'
  resp = GET(url, query = list(key = api_key, 
                               country = country))
  parsed = fromJSON(rawToChar(resp$content))
  
  return (as.list(parsed$data)$state)
}
```


## Function to get cities

Function to get `cities` based on the provided `country` and `state`. Cities returned as a list.


```{r}
get_cities <- function(country, state, api_key = key) {
  url = 'http://api.airvisual.com/v2/cities'
  
  resp = GET(url, query = list(key = api_key, 
                               country = country,
                               state = state))
  parsed = fromJSON(rawToChar(resp$content))
  return (as.list(parsed$data)$city)
}
```


## Function get air quality and weather metrics based on the country, state and city

Function returns air quality based on the `country`, `state` and `city`. 

If `city` is not provided, random sample of the size `sample_size` will be chosen from the cities which belong to the `state`. 


If `weather_metrics = TRUE` in addition to air quality metrics, weather metrics like, `temperature`, `humidity`, `wind speed` and `atmospheric pressure` will be returned. 

Function returns data frame in `tibble` format with following columns:

  - `country`
  - `state`
  - `city`
  - `aqius` - AQI value based on US EPA standard
  - `aqicn` - AQI value based on China MEP standard

If  `weather_metrics = TRUE` in addition to the columns above, function also returns:


  - `temp_cels` - temperature in Celsius
  - `humidity` - humidity %
  - `wind_speed` - wind speed (m/s)
  - `atm_pressure` - atmospheric pressure in hPa


```{r}
get_air_quality_per_city <- function(country, 
                                     state, 
                                     city = NULL,
                                     weather_metrics = TRUE,
                                     sample_size = 5,
                                     api_key = key) {
  
  results = tibble()
  url = 'http://api.airvisual.com/v2/city'
 
  if (is.null(city)) {
    cities = get_cities(country, state)
  
    if (length(cities) > sample_size) {
      set.seed(123)
      cities = sample(cities, size = sample_size)
    }
    
    for (city in cities) {
      Sys.sleep(15) # required to avoid error - too many requests
      resp = GET(url, query = list(key = key, 
                                   country = country,
                                   state = state,
                                   city = city))
      parsed = fromJSON(rawToChar(resp$content))
      
      if (weather_metrics == TRUE) {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country,
                           temp_cels = parsed$data$current$weather$tp,
                           humidity = parsed$data$current$weather$hu, 
                           wind_speed = parsed$data$current$weather$ws,
                           atm_pressure = parsed$data$current$weather$pr)
      } else {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country)
      }
      results <- bind_rows(results, subset_df)
    
  }
    } else {
      resp = GET(url, query = list(key = key, 
                                   country = country,
                                   state = state,
                                   city = city))
      parsed = fromJSON(rawToChar(resp$content))
      if (weather_metrics == TRUE) {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country,
                           temp_cels = parsed$data$current$weather$tp,
                           humidity = parsed$data$current$weather$hu, 
                           wind_speed = parsed$data$current$weather$ws,
                           atm_pressure = parsed$data$current$weather$pr)
      } else {
        subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                           aqicn = parsed$data$current$pollution$aqicn, 
                           city = parsed$data$city,
                           state = parsed$data$state,
                           country = parsed$data$country)
      }
      results <- bind_rows(results, subset_df)
  }
  return (results)
}
```


## Function to get air quality and weather metrics for country

Function returns air quality and weather metrics based on the country name. Random sample of the size `sample_size` is selected from from the states which belongs to the country. A random city is selected for each state, and metrics are then retrieved.


Function returns data frame in `tibble` format with following columns:

  - `country`
  - `state`
  - `city`
  - `aqius` - AQI value based on US EPA standard
  - `aqicn` - AQI value based on China MEP standard
  - `temp_cels` - temperature in Celsius
  - `humidity` - humidity %
  - `wind_speed` - wind speed (m/s)
  - `atm_pressure` - atmospheric pressure in hPa


```{r}
get_air_quality_per_country <- function(country, 
                                        weather_metrics = TRUE,
                                        sample_size = 5,
                                        api_key = key) {
  
  results = tibble()
  url = 'http://api.airvisual.com/v2/city'
  
  
  states = get_states(country, api_key = key)
  
  set.seed(2)
  states = sample(states, size = sample_size)
  
  
  for (state in states) {
    Sys.sleep(15) # required to avoid error - too many requests
    cities = get_cities(country, state, api_key = key)
    set.seed(4)
    city = sample(cities, size = 1)
    
    Sys.sleep(15) # required to avoid error - too many requests
    resp = GET(url, query = list(key = api_key, 
                                 country = country,
                                 state = state,
                                 city = city))
    parsed = fromJSON(rawToChar(resp$content))
    subset_df = tibble(aqius = parsed$data$current$pollution$aqius, 
                       aqicn = parsed$data$current$pollution$aqicn, 
                       city = parsed$data$city,
                       state = parsed$data$state,
                       country = parsed$data$country,
                       temp_cels = parsed$data$current$weather$tp,
                       humidity = parsed$data$current$weather$hu, 
                       wind_speed = parsed$data$current$weather$ws,
                       atm_pressure = parsed$data$current$weather$pr)
    results <- bind_rows(results, subset_df)
  }
  return (results)
}
```


# Exploratory Data Analysis

EDA analysis will be conducted at the country level and, for the United States, at the state level.

## EDA for countries


Let's look at air quality and weather metrics for different countries across the globe. Chosen countries: **China**, **USA**, **Poland**, **Australia** and **Norway**.

*Note: Sample size and set of countries chosen to be small due to API limitations and time required to render the whole document* 

```{r, cache=TRUE}
sample = 5
air_usa = get_air_quality_per_country(country = 'USA', sample_size = sample)
air_china = get_air_quality_per_country(country = 'China', sample_size = sample)
air_poland = get_air_quality_per_country(country = 'Poland', sample_size = sample)
air_australia = get_air_quality_per_country(country = 'Australia', sample_size = sample)
air_norway = get_air_quality_per_country(country = 'Norway', sample_size = sample)
```

Combine all obtained data frames together in one data frame.

```{r}
all = bind_rows(air_australia, air_poland, air_china, air_usa, air_norway)
```

Let's look at data frame with combined records:
```{r}
all
```

Now, we need to create categorical variables for `aqius`, `aqicn` and `humidity`. 

If `aqius` value is less than **50**, the air quality is considered to be `good`. If `aqius` value is greater or equal **50** and less or equal **100**,  the air quality is considered to be `moderate`. If `aquis` value is greater than **100**, the air quality is considered `unhealthy`.

If `aqicn` value is less than **50**, the air quality is considered to be `excellent`, if `aqicn` value is greater or equal **50** and less or equal **100**, the air quality is considered to be `good`.  If `aqicn` value is greater than **100**, the air quality is considered `polluted`.

If `humidity` level is less or equal **40%**, it is considered `low`. If `humidity` level is greater than **40%** and less than or equal to **60%**, it is considered `normal`. If `humidity` is greater than **60%**, it is considered `high`.

Creating new categorical variables based on the rules outlined above and displaying updated data frame.


```{r}
all <- all %>%
  mutate(aqius_category = case_when(
    aqius < 50 ~ "good",
    aqius >= 50 & aqius <= 100 ~ "moderate",
    aqius > 100 ~ "unhealthy"
  )) %>%
  mutate(aqicn_category = case_when(
    aqicn < 50 ~ "excellent",
    aqicn >= 50 & aqicn <= 100 ~ "good",
    aqicn > 100 ~ "polluted"
  )) %>%
  mutate(humidity_category = case_when(
    humidity <= 40 ~ 'low',
    humidity > 40 & humidity <= 60 ~ 'normal',
    humidity > 60 ~ 'high'
  ))
 all 
```


Create 2-way contingency table for `aqius_category` vs `country`.

```{r}
table(all$aqius_category, all$country)
```
As it could be seen from the table above, `China` has at least one city with `unhealthy` air based on `aqius` values. All sampled cities for `Australia` and `Norway` have `good` air quality. 

Create 2-way contingency table for `aqicn_category` vs `country`.

```{r}
table(all$aqicn_category, all$country)
```

Based on the `aqicn` values, all selected cities for all countries, except `China` have `excellent` air quality.

Create 2-way contingency table for `aqius_category` vs `humidity_category`.

```{r}
table(all$aqius_category, all$humidity_category)
```


Create 2-way contingency table for `aqicn_category` vs `humidity_category`.

```{r}
table(all$aqicn_category, all$humidity_category)
```

According to the two tables above, the highest number of observations for `excellent` and `good` air quality is associated with high humidity levels.


Now, let's group data by `country` and calculate `aqius` mean, `aqicn` mean and standard deviations for each group.

```{r}
result <- all %>%
  group_by(country) %>%
  summarize(mean_aqius = mean(aqius), 
            mean_aqicn = mean(aqicn), 
            sd_aquis = sd(aqius),
            sd_aqicn = sd(aqicn)) 
result
```
Based on the values above, `China` has the highest values for `aqius` and `aqicn`, while `Australia` has the lowest values for `aqius` and `aqicn`. 


Visualizing means for `aqius`.


```{r}
ggplot(all, aes(x = country, y = aqius)) +
  geom_bar(stat = "summary", 
           position = 'dodge', 
           fun = 'mean',  
           fill = "blue", 
           alpha = 0.5) +
  labs(title = "Means of AQIUS per country", x = "Country", y = "Mean of AQIUS")
```

Visualizing means for `aqicn`.

```{r}
ggplot(all, aes(x = country, y = aqicn)) +
  geom_bar(stat = "summary", 
           position = 'dodge', 
           fun = 'mean',  
           fill = "red", 
           alpha = 0.5) +
  labs(title = "Means of AQICN per country", x = "Country", y = "Mean of AQICN")
```

The graphs above shows means for the `aqius` and `aqicn` for the selected coutries. `China` has the highest values for `aqius` and `aqicn`, while `Australia` has the lowest values for `aqius` and `aqicn`.

Let's look at `aqius` distribution for each country. Creating box plot for `aqius` per `country`.

```{r}
ggplot(all, aes(x = country, y = aqius, fill = country)) +
  geom_boxplot() +
  labs(title = "AQIUS values distribution per country", x = "Country", y = "AQIUS") +
  theme_minimal()
```

According to the graph above, China has the largest median for `aqius` and the largest standard deviation among the other selected countries.

Creating heatmap for numerical variables

```{r}
data <- as.matrix(all %>% select(aqius, aqicn, temp_cels, humidity, wind_speed, atm_pressure))
heatmap(data,
        Rowv=NA,
        Colv=NA, 
        labCol=colnames(data), 
        col= colorRampPalette(brewer.pal(8, "Oranges"))(25), 
        scale="column",
        main="Heatmap for numerical variables")
legend(x="bottomright", legend=c("min", "ave", "max"), 
     fill=colorRampPalette(brewer.pal(8, "Oranges"))(3))
```

Let's looks at correlation matrix for all numerical variables.

```{r}
cor(data)
```

From the matrix above, it could be seen that there are some positive linear correlation between `aqius` and `atm_pressure`, `aqius` and `temp_cels`, `aqicn` and `temp_cels`.  There are some negative linear correlation between `aqius` and `wind_speed`, `aqius` and `humidity`, `aqicn` and `humidity`, `aqicn` and `wind_speed`.


Let's visualize obtained results for correlation:

```{r}
corrplot(cor(data), 
         type="upper", 
         method = "number",
         tl.pos = "lt")
```

Let's check if there are any linear dependency between current temperature and observed `aquis` values.

```{r}
ggplot(all, aes(x = temp_cels, y = aqius)) +
  geom_point() +          
  geom_smooth(method = "lm",
              formula = y ~ x,
              se = TRUE) +
  labs(title = "Temperature vs AQIUS",
       x = "Temperature, Celsius",
       y = "AQIUS")  
```

As, it could be seen from the graph above, there are might be some relationship between observed `temperature` and observed `aquis` values. The higher the temperature, the higher the `aquis` values.



## EDA for the US states

Now, let's look at several states across `USA` and compare air quality in the context . Chosen states: *California*, *Colorado*, *Minnesota*, *Florida* and *Michigan*. Sample size for each state is equal to `5`.


Reading data for each state using function `get_air_quality_per_city`:

```{r, cache=TRUE}
sample = 5
air_california = get_air_quality_per_city(country = 'USA', state = 'California', sample_size = sample)
air_colorado = get_air_quality_per_city(country = 'USA', state = 'Colorado', sample_size = sample)
air_minnesota = get_air_quality_per_city(country = 'USA', state = 'Minnesota', sample_size = sample)
air_florida = get_air_quality_per_city(country = 'USA', state = 'Florida', sample_size = sample)
air_michigan = get_air_quality_per_city(country = 'USA', state = 'Michigan', sample_size = sample)
```

Combine results in one data frame:

```{r}
all_per_state = bind_rows(air_michigan, air_florida, air_minnesota, air_colorado, air_california)
all_per_state
```


Now, let's use the same rules as outlined above, to create categorical variables `aqius_category`, `aqicn_category` and `humidity_category` and display updated data frame.

```{r}
all_per_state <- all_per_state %>%
  mutate(aqius_category = case_when(
    aqius < 50 ~ "good",
    aqius >= 50 & aqius <= 100 ~ "moderate",
    aqius > 100 ~ "unhealthy"
  )) %>%
  mutate(aqicn_category = case_when(
    aqius < 50 ~ "excellent",
    aqius >= 50 & aqius <= 100 ~ "good",
    aqius > 100 ~ "polluted"
  )) %>%
  mutate(humidity_category = case_when(
    humidity <= 40 ~ 'low',
    humidity > 40 & humidity <= 60 ~ 'normal',
    humidity > 60 ~ 'high'
  ))
all_per_state 
```


Create 2-way contingency table for `aqius_category` vs `state`.


```{r}
table(all_per_state$aqius_category, all_per_state$state)
```

Based on the table above, `Florida`, `Colorado` and `Michigan` have all selected states with `good` air quality. `California` and `Minnesota` have 1 state out of 5 states with `moderate` air quality.

Create 2-way contingency table for `aqicn_category` vs `state`.

```{r}
table(all_per_state$aqicn_category, all_per_state$state)
```


Based on the table above, `Florida`, `Colorado` and `Michigan` have all selected states with `excellent` air quality. `California` and `Minnesota` have 1 state out of 5 states with `good` air quality.


Create 2-way contingency table for `aqius_category` vs `humidity_category`.


```{r}
table(all_per_state$aqius_category, all_per_state$humidity_category)
```

Most of the observations with `good` air quality are from observations with `high` humidity. 

Create 2-way contingency table for `aqicn_category` vs `humidity_category`.


```{r}
table(all_per_state$aqicn_category, all_per_state$humidity_category)
```
Most of the observations with `excellent` air quality are from observations with `high` humidity. 

Let's looks at correlation matrix for all numerical variables.

```{r}
data <- as.matrix(all_per_state %>% select(aqius, aqicn, temp_cels, humidity, wind_speed, atm_pressure))
cor(data)
```

From the matrix above, it could be seen that there are some positive linear correlation between  `aqius` and `temp_cels`. There are some negative correlation between `aqius` and `wind_speed`, `aqicn` and `wind_speed`.


Let's visualize obtained results for correlation:

```{r}
corrplot(cor(data), 
         type="upper", 
         method = "number",
         tl.pos = "lt")
```

Now, let's group by `state` and calculate `aqius` mean, `aqicn` mean and standard deviations for each group.

```{r}
result <- all_per_state %>%
  group_by(state) %>%
  summarize(mean_aqius = mean(aqius), 
            mean_aqicn = mean(aqicn), 
            sd_aquis = sd(aqius),
            sd_aqicn = sd(aqicn)) 
result
```


According to the table above, `Florida` has the highest mean values for `aqius` and `aqicn` and the highest standard deviations for both metrics.

Let's look at linear relationship between `wind speed` and observed `aquis` values.


```{r}
ggplot(all_per_state, aes(x = wind_speed, y = aqius)) +
  geom_point() +                 
  geom_smooth(method = "lm", 
              formula = y ~ x,
              se = TRUE) + 
  labs(title = "Temperature vs AQIUS", 
       x = "Wind speed",
       y = "AQIUS")  
```


As it could be seen from the graph, the stronger the wind, the lower the `aqius` values. 


Let's check if there are any linear dependency between temperature and observed `aquis` values.  

```{r}
ggplot(all_per_state, aes(x = temp_cels, y = aqius)) +
  geom_point() + 
  geom_smooth(method = "lm", 
              formula = y ~ x,
              se = TRUE) + 
  labs(title = "Temperature vs AQIUS", 
       x = "Temperature", 
       y = "AQIUS")  
```

As it could be seen from the graph above, the higher the temperature, the higher `aqius` values.


Let's check if there are any linear dependency between `humidity` level and observed `aquis` values. 

```{r}
ggplot(all_per_state, aes(x = humidity, y = aqius)) +
  geom_point() + 
  geom_smooth(method = "lm",
              formula = y ~ x, 
              se = TRUE) + 
  labs(title = "Humidity vs AQIUS", 
       x = "Humidity",
       y = "AQIUS") 
```

As it could be seen from the graph above, there is no linear relationship between `humidity` levels and `aqius` values.


Let's check if there are any linear dependency between atmospheric pressure and observed `aquis` values. 

```{r}
ggplot(all_per_state, aes(x = atm_pressure, y = aqius)) +
  geom_point() + 
  geom_smooth(method = "lm",
              formula = y ~ x, 
              se = TRUE) + 
  labs(title = "Atmospheric pressure vs AQIUS", 
       x = "Atmospheric pressure in hPa",
       y = "AQIUS") 
```

According to the graph above, there is no relationship (or very minor) between atmospheric pressure and the `aqius`.


Let's look at `aqius` distribution for each `state`. Creating box plot for `aqius` per `state`.

```{r}
ggplot(all_per_state, aes(x = state, y = aqius, fill = state)) +
  geom_boxplot() +
  labs(title = "AQIUS per state", x = "State", y = "AQIUS") +
  theme_minimal()
```

According to the graph above, Minnesota has the highest median value for `aqius` among selected states.

# Summary

In this project, functions were created to interact with the **IQAir APIs**. These functions allow us to retrieve data on air quality and weather metrics for countries, states, and cities. Due to limitations on free account, small samples were obtained for several countries worldwide and several states within the United States. For both data sets, exploratory data analysis was conducted, revealing the following insights:

  - air quality tends to decrease with rising temperature
  - air quality tends to increase with increasing wind speed
  - `Minnesota` has the poorest air quality among the selected states, while `China` exhibits the worst air quality among the selected countries.
  
**IQAir APIs** are more suitable to retrieve granular information per city. In order to retrieve large sample of the data, many API calls should be done. 
  


